formula1

### 2020-06-02 inst/f1points.Rmd
more work, proper comparison between points and likelihood ranking

The Bradley-Terry model \citep{bradley1952} assigns non-negative
strengths $p_1,\ldots, p_n$ to each of $n$ competitors in such a way
that the probability of $i$ beating $j\neq i$ in pairwise competition
is $\frac{p_i}{p_i+p_j}$; it is conventional to normalize so that
$\sum p_i=1$.  Further, we use a generalization due to
\citet{luce1959}, in which the probability of competitor $i$ winning
in a field of $j=1,\ldots, n$ is $\frac{p_i}{p_1+\cdots +p_n}$.
Noting that there is information in the whole of the finishing order,
not just the first across the line, we can follow \cite{plackett1975}
and consider the runner-up to be the winner among the remaining
competitors; and so on down the finishing order.  Without loss of
generality, if the order of finishing were $1,2,3,4$, then a suitable
\citeauthor{plackett1975} likelihood function would be

\begin{equation}\label{competitors_1_to_4_likelihood}
\frac{p_1}{p_1+p_2+p_3+p_4}\cdot
\frac{p_2}{p_2+p_3+p_4}\cdot
\frac{p_3}{p_3+p_4}\cdot
\frac{p_4}{p_4}\cdot
\end{equation}

We now use a technique due to \cite{hankin2010,hankin2017} and
introduce fictional (reified) entities whose nonzero Bradley-Terry
strength helps certain competitors or sets of competitors.  The
canonical example would be the home-ground advantage in association
football.  If players (teams) $1,2$ with strengths $p_1,p_2$ compete,
and if our observation were $a$ home wins and $b$ away wins for team
$1$, and $c$ home wins and $d$ away wins for team $2$, then a suitable
likelihood function would be

@@

We see the that the top two players, Hamilton and Vettel
of roughly equal strength and indeed a likelihood ratio test
[`samep.test()`] does not reject the null of equality ($p=0.09$).

@@

```{r,label=likevspoints,cache=TRUE,echo=FALSE}
jj <- read.table("formula1_2017.txt",header=TRUE)
points <- jj[,21]
names(points) <- rownames(jj)

mL2017 <- maxp(H)
m <- mL2017   # maxp(H)
ox <- round(rank(-points[seq_len(top)]))  # negative because number 1 = largest number of points
oy <- rank(-m)                            # negative because number 1 = largest strength
par(pty='s')        # square plot
plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
    xlab="official order",ylab="my order",main='Formula 1, 2017 season')
par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray') }
par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
abline(0,1)
```

From the above, we see that

>>>>

The Bradley-Terry model \citep{bradley1952} assigns non-negative
strengths $p_1,\ldots, p_n$ to each of $n$ competitors in such a way
that the probability of $i$ beating $j\neq i$ in pairwise competition
is $\frac{p_i}{p_i+p_j}$; it is conventional to normalize so that
$\sum p_i=1$.  Further, we use a generalization due to
\citet{luce1959}, in which the probability of competitor $i$ winning
in a field of $j=1,\ldots, n$ is $\frac{p_i}{p_1+\cdots +p_n}$.
Noting that there is information in the whole of the finishing order,
not just the first across the line, we can follow \cite{plackett1975}
and consider the runner-up to be the winner among the remaining
competitors; and so on down the finishing order.  Without loss of
generality, if the order of finishing were $1,2,3,4,5$, then a suitable
\citeauthor{plackett1975} likelihood function would be

\begin{equation}\label{competitors_1_to_5_likelihood}
\frac{p_1}{p_1+p_2+p_3+p_4+p_5}\cdot
\frac{p_2}{p_2+p_3+p_4+p_5}\cdot
\frac{p_3}{p_3+p_4+p_5}\cdot
\frac{p_4}{p_4+p_5}\cdot
\frac{p_5}{p_5}
\end{equation}

and a slight generalization of this would allow the incorporation of
non-finishers (DNF etc).  If, say, competitors 4 and 5 did not finish,
we would have

\begin{equation}\label{competitors_1_to_3_only_finished}
\frac{p_1}{p_1+p_2+p_3+p_4+p_5}\cdot
\frac{p_2}{p_2+p_3+p_4+p_5}\cdot
\frac{p_3}{p_3+p_4+p_5}
\end{equation}

(observe how this likelihood function, while informative about
$p_4+p_5$, is uninformative about $p_4\left|p_4+p_5\right.$).  We now
use a technique due to \cite{hankin2010,hankin2017} and introduce
fictional (reified) entities whose nonzero Bradley-Terry strength
helps certain competitors or sets of competitors.  The canonical
example would be the home-ground advantage in association football.
If players (teams) $1,2$ with strengths $p_1,p_2$ compete, and if our
observation were $a$ home wins and $b$ away wins for team $1$, and $c$
home wins and $d$ away wins for team $2$, then a suitable likelihood
function would be

@@

We see the Hamilton has the largest estimated strength, although it
should be noted that a likelihood ratio test [`samep.test()`] does not
reject the null of $p_\mathrm{Hamilton}=p_\mathrm{Vettel}$ ($p=0.09$).

@@

```{r,label=calcpoints}
p <- c(25, 18, 15, 12, 10, 8, 6, 4, 2, 1, 0, 0)
dp <- ordertable2points(as.ordertable(F1_table_2017[seq_len(top),]),p)
dp
dp <- dp[order(-dp)]
dp[] <- seq_along(dp)
```
now likelihood

```{r,label=calclike}
m <- maxp(H)
m <- sort(m,decreasing=TRUE)
m[] <- seq_along(m)
```

now ranks:

```{r}
jj <- outer(names(m),names(dp),"==")
ox <- dp
oy <- m[which(jj,arr.ind=TRUE)[,1]]
```

and plot:

```{r,label=likevspoints,cache=TRUE,echo=FALSE}
par(pty='s')        # square plot
plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
    xlab="official order",ylab="my order",main='Formula 1, 2017 season')
par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray') }
par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
abline(0,1)
```

From the above, we see that the first four drivers (Hamilton, Vettel,
Raikkonen, Bottas) are respectively first, second, third and fourth
according to both ranking procedures.  The lower-ranked players,
having estimated strengths of less than 0.03, are more scattered as
one might expect.

We can use this observation to assess a given points system: using
Monte-Carlo techniques we can generate an order table \emph{in
silico}, using estimated driver strengths from the 2017 table, and
then compare rankings generated by estimated \citeauthor{plackett1975}
strengths with rankings generated by the points system.

# Monte Carlo assumptions

blah blah
